# =============================================================================
# 1_timeseriesAnalysis.Rmd
# Purpose : Main analysis. Produces (1) CD4-at-diagnosis "rainbow" plots,
#           (2) joinpoint / segmented trend analysis of HIV notifications,
#           (3) PHN-level notification rates, heatmaps, and geographic
#           inequality (Gini/Lorenz), for first-ever-diagnosed-in-Australia
#           (FED) and previously-diagnosed-overseas (PDO) groups, 2001-2024.
# Run order: 0_Setupmodel.Rmd  ->  this file. Sources three helper scripts
#           (Joinpoint_analysis.R, Rainbow_plot_and_stats.R, PHN_timeseries.R).
# INPUTS  (place in projects/<project>/data/ unless noted):
#   - hiv_data_cleaned_YYYYMMDD.csv  : de-identified line-list of HIV notifications.
#   - AU_pop_size_2001_2024.csv      : national population by year.
#   - SA2_pop_size_2001_2024.csv     : SA2-level population by year (ABS).
#   - phn_sa2_20250901.csv           : SA2-to-PHN concordance.
# OUTPUTS : CSV tables in projects/<project>/output/, figures in .../figures/.
# =============================================================================

```{r Initialise}
##setwd()

#######################r initialization####################
# Clear workspace
rm(list=ls()) 

# Load libraries
# Install required packages if not already installed
# install.packages(c( "spdep", "tmap", "epitools", "janitor", "lubridate"))
library(readr)        # For reading CSV files
library(tidyverse)    # For data manipulation and plotting
library(sf)           # For reading and handling spatial data
library(spdep)        # For spatial analysis, including Getis-Ord Gi*
library(epitools)     # For potential rate calculations (if needed)
library(janitor)      # For data cleaning
library(lubridate)    # For date manipulation
#library(knitr)        # For creating tables
library(kableExtra)   # For table formatting
library(dplyr)         # For data manipulation
library(ineq)

# Source to current directory and set working directory
basePath <- getwd()

# Restart R and set working directory to source file location. 

##r User inputs
##select project
selectedproject<- "2000_2024_changing_point"

# Various directories
projectFolder <- file.path(basePath,"projects", selectedproject)
codeFolder <- file.path(basePath,"code")
dataFolder <- file.path(basePath,"projects", selectedproject,"data")
resultsFolder <- file.path(basePath,"projects", selectedproject, "output")
figFolder <- file.path(basePath,"projects", selectedproject,"figures")

# NOTE: absolute path to the restricted-access source data on the analyst's
# machine. Not reproducible off this machine by design — the line-list is not
# public. Users: point HIVDataFolder to your own local copy obtained
# under the relevant data agreement. Nothing downstream writes raw data.
# HIV data folder - for sourcing latest notification cleaning files
HIVDataFolder <- file.path("C:", "onedrive", "OneDrive - UNSW", "HIV Trends Across Space and Time", 
  "General")

# Source
source(file.path(codeFolder, "Joinpoint_analysis.R"))
source(file.path(codeFolder, "Rainbow_plot_and_stats.R"))
source(file.path(codeFolder, "PHN_timeseries.R"))



```



```{r Input}
# Specify project input files and load them
hivdataFile<- file.path(HIVDataFolder, "hiv_data_cleaned_20250731.csv")
# Read the CSV data
hiv_data_all <- read_csv(hivdataFile, col_types = cols(
  month_hiv_diag = col_character(),
  age_hiv_notif = col_double(),
  sex = col_character(),
  gender = col_character(),
  region_birth = col_character(),
  au_birth = col_character(),
  lang_home = col_character(),
  year_arrival_aus = col_double(),
  cd4_count_diag = col_double(),
  late_hiv_diag_flag = col_logical(), 
  newly_acq_hiv_flag = col_logical(),
  hiv_exposure_risk = col_character(),
  place_acq = col_character(),
  state_res = col_character(),
  phn_res = col_character(),
  phn_name = col_character(),
  previ_diag_overseas = col_logical(),
  id = col_double(),
  year_diag = col_double()
))

# Derived clinical stage flags. "Advanced HIV at diagnosis" = CD4 < 200
# cells/mm3 AND meeting the late-diagnosis criterion; NA propagated when late-
# diagnosis status is unknown.
hiv_data_all <- hiv_data_all %>%
  mutate(
    advanced_hiv_diag_flag = case_when(
      # If late_hiv_diag_flag is NA, return NA
       is.na(late_hiv_diag_flag) ~ NA,
      # If both conditions are met (cd4 < 200 AND late diagnosis), return TRUE
      cd4_count_diag < 200 & late_hiv_diag_flag == TRUE ~ TRUE,
      # All other cases return FALSE
      TRUE ~ FALSE
    )
  )

# Same construction as above but for CD4 < 50 cells/mm3.
hiv_data_all <- hiv_data_all %>%
  mutate(
    cd4_50_hiv_diag_flag = case_when(
      # If late_hiv_diag_flag is NA, return NA
      is.na(late_hiv_diag_flag)  ~ NA,
      # If both conditions are met (cd4 < 50 AND late diagnosis), return TRUE
      cd4_count_diag < 50 & late_hiv_diag_flag == TRUE ~ TRUE,
      # All other cases return FALSE
      TRUE ~ FALSE
    )
  )
# Analytic population: adults (>=18y at diagnosis). Records with missing age are
# RETAINED (is.na(...) | ...) to avoid dropping notifications with unknown age;
# see missingness chunk for the extent of missing age.
hiv_data <- hiv_data_all %>%
  filter(
    is.na(age_hiv_notif) | age_hiv_notif >= 18
  )

# Check data structure
str(hiv_data)

```



```{r rainbow plot}

####Run the function to generate the rainbow plot####
# All data, default settings
rainbow_all <- create_cd4_rainbow_plot(
  data = hiv_data,
  start_year = 2001,
  end_year = 2024,
  title = "",
  subtitle = "A. Proportion of HIV notifications",
  show_n = FALSE,  # Do not show sample size on x-axis
  previ_diag_overseas = FALSE,
  dash_line = FALSE
)
rainbow_all$summary_data_complete
rainbow_all_percentage_plot <- rainbow_all$p + theme(legend.position = "none")
print(rainbow_all)

if (!file.exists(file.path(resultsFolder, "rainbow_plot_prop_20260504.csv"))) {
write.csv(rainbow_all$summary_data_complete, file.path(resultsFolder, "rainbow_plot_prop_20260504.csv"), row.names=FALSE)
}


####save figure####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 40, 
    height = 20, units = "cm", dpi = 600,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}

# If figure not already created make sure it is not overwritten
if (!file.exists(file.path(figFolder, "rainbow_all_noPOD_20250926.png"))) {
    SaveFigure(figFolder,"rainbow_all_noPOD_20250926", rainbow_all$p)
}

####Run the function to generate the rainbow plot for absolute number####
absolute_rainbow_all <- create_cd4_absolute_rainbow_plot(
  data = hiv_data,
  start_year = 2001,
  end_year = 2024,
  title = "",
  subtitle = "B. Monthly average number of HIV notifications",
  monthly_average = TRUE, 
  show_n = FALSE,  # Do not show sample size on x-axis
  previ_diag_overseas = FALSE,  # Do not include previously diagnosed overseas
  dash_line = FALSE
)
print(absolute_rainbow_all)

rainbow_all_percentage_plot <- rainbow_all$p + theme(legend.position = "none")
rainbow_all_number_plot <- absolute_rainbow_all
# Combine side by side (or adjust ncol/nrow as needed)
library(patchwork)
combined_number_percentage_plot <- 
  (rainbow_all_percentage_plot + rainbow_all_number_plot) +
  plot_layout(heights = c(1, 1))

####save figure####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 30, 
    height = 20, units = "cm", dpi = 600,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}


# If figure not already created make sure it is not overwritten
if (!file.exists(file.path(figFolder, "perNnumber_rainbow_all_noPDO_202605041.png"))) {
    SaveFigure(figFolder,"perNnumber_rainbow_all_noPDO_20260504", combined_number_percentage_plot)
}

#### cd4 trend ####
# Restrict to fed + non-missing CD4
hiv_cd4_fed <- hiv_data %>%
  filter(
    year_diag >= 2001,
    year_diag <= 2024,
    !is.na(year_diag),
    !is.na(cd4_count_diag),
    previ_diag_overseas == 0
  )

# Annual mean & median CD4
annual_cd4_summary <- hiv_cd4_fed %>%
  group_by(year_diag) %>%
  summarise(
    mean_cd4   = mean(cd4_count_diag, na.rm = TRUE),
    median_cd4 = median(cd4_count_diag, na.rm = TRUE),
    n          = n(),              # optional: number of people each year
    .groups    = "drop"
  ) %>%
  rename(year = year_diag)

cd4_trends <- list(
  mean   = lm(mean_cd4 ~ year, data = annual_cd4_summary),
  median = lm(median_cd4 ~ year, data = annual_cd4_summary)
) %>%
  lapply(tidy) %>%
  bind_rows(.id = "outcome")

cd4_time_plot <- ggplot(annual_cd4_summary, aes(x = year, y = median_cd4)) +
  geom_line(color = "#B696B6", linewidth = 1) +
  geom_point(color = "#B696B6", size = 2) +
  scale_x_continuous(breaks = seq(2001, 2024, 3)) +
  scale_y_continuous(limits = c(0, 600),
                       expand = c(0, 0)) +   # y-axis starts at 0
  labs(
    title = "",
    y = "CD4 count (cells/mm^3)",
    x = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),     # remove major gridlines
    panel.grid.minor = element_blank(),     # remove minor gridlines
    axis.line = element_line(color = "black"),  # add black x & y axes
    axis.ticks = element_line(color = "black"), # add tick marks
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", size = 12)
  )
cd4_time_plot

# Save figure
if (!file.exists(file.path(figFolder, "cd4_time_plot_202511191.png"))) {
    SaveFigure(figFolder,"cd4_time_plot_20251119", cd4_time_plot,
               width = 20, height = 10)
}


```




```{r Trend Analysis Functions}

library(segmented)
library(MASS)


#### # Example usage of the functions above ####
#### All notifications ####
segmented_results <- analyze_segmented_trend_complete(
  data = hiv_data,
  start_year = 2001,
  end_year = 2024,
  count_period = "monthly",
  include_overseas = FALSE,
# FED = "first-ever diagnosed in Australia" (previously_diagnosed_overseas == 0).
# PDO (below) = previously diagnosed overseas (== 1). These two groups are
# analysed separately.
  family_type  = "NB",
# Negative binomial rather than Poisson: monthly notification counts are
# over-dispersed (variance > mean).
  pc_period    = "monthly",
  num_changes = 5,
  main_title = "",
  sub_title = "A. HIV notifications with a first ever diagnosis in Australia",
  show_overall = FALSE,  # Don't show overall trend
  show_segment = TRUE,   # Don't show segmented trends,
  y_upper = 150,    # Set y-axis upper limit
  y_break = 50      # Set y-axis breaks every 50
)


segmented_results$segment_results
segmented_results$segment_plot
segmented_results$overall_results
segmented_results$bic_candidates
segmented_results$change_points
segmented_results$wald_adjacent_slopes


#### Previously diagnosed overseas ####
PDO_segmented_results <- analyze_segmented_trend_complete(
  data = hiv_data,
  start_year = 2001,
  end_year = 2024,
  count_period = "monthly",
  include_overseas = TRUE,
  family_type = "NB",
  pc_period    = "monthly",
  main_title = "",
  sub_title = "B. HIV notifications previously diagnosed overseas",
  num_changes = 5,
  show_overall = FALSE,  # Don't show overall trend
  show_segment = TRUE   # Don't show segmented trends
)

PDO_segmented_results$segment_results
PDO_segmented_results$segment_plot
PDO_segmented_results$overall_results
PDO_segmented_results$bic_candidates
PDO_segmented_results$change_points
PDO_segmented_results$wald_adjacent_slopes

#### Combined plot (first time diagnosed in Australia vs previously diagnosed overseas) ####
library(grid)
library(gtable)
library(patchwork)
FED_PDO_segment_plot <- (
  wrap_plots(
    segmented_results$segment_plot,
    PDO_segmented_results$segment_plot,
    nrow = 2
  ) + plot_layout(guides = "collect")
) & theme(legend.position = "bottom")

#### Save figures ####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 25, # 30 , 15 for figure 1 in paper, 25,12.5 for 2 combined F in slides, 20, 10for 1 in slides
    height = 12.5, units = "cm", dpi = 600,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}
# If figure not already created make sure it is not overwritten
if (!file.exists(file.path(figFolder, "FED_PDO_segment_plot_2001_2024_20260707.png"))) {
  SaveFigure(figFolder,"FED_PDO_segment_plot_2001_2024_20260707", FED_PDO_segment_plot)
}

#### Export results to CSV files ####
# Writes the FED + PDO joinpoint results (per-segment slopes, change points,
# BIC candidates, adjacent-slope Wald tests, AAPC) to output/ as combined_*.csv.
# Existing files are never overwritten (guarded by file.exists()).
# Function to combine results across populations
combine_segmented_results <- function(results_list, population_labels) {
  
  # 1. Combine segment_results
  segment_combined <- do.call(rbind, lapply(seq_along(results_list), function(i) {
    df <- results_list[[i]]$segment_results
    if (!is.null(df) && nrow(df) > 0) {
      df$Population <- population_labels[i]
      df
    } else {
      NULL
    }
  }))
  
  # Reorder columns to put Population first
  if (!is.null(segment_combined)) {
    segment_combined <- segment_combined[, c("Population", setdiff(names(segment_combined), "Population"))]
  }
  
  # 2. Combine bic_candidates
  bic_combined <- do.call(rbind, lapply(seq_along(results_list), function(i) {
    df <- results_list[[i]]$bic_candidates
    if (!is.null(df) && nrow(df) > 0) {
      df$Population <- population_labels[i]
      df
    } else {
      NULL
    }
  }))
  
  if (!is.null(bic_combined)) {
    bic_combined <- bic_combined[, c("Population", setdiff(names(bic_combined), "Population"))]
  }
  
  # 3. Combine change_points
  change_points_combined <- do.call(rbind, lapply(seq_along(results_list), function(i) {
    df <- results_list[[i]]$change_points
    if (!is.null(df) && nrow(df) > 0) {
      df$Population <- population_labels[i]
      df
    } else {
      NULL
    }
  }))
  
  if (!is.null(change_points_combined)) {
    change_points_combined <- change_points_combined[, c("Population", setdiff(names(change_points_combined), "Population"))]
  }
  
  # 4. Combine wald_adjacent_slopes
  wald_combined <- do.call(rbind, lapply(seq_along(results_list), function(i) {
    df <- results_list[[i]]$wald_adjacent_slopes
    if (!is.null(df) && nrow(df) > 0) {
      df$Population <- population_labels[i]
      df
    } else {
      NULL
    }
  }))
  
  if (!is.null(wald_combined)) {
    wald_combined <- wald_combined[, c("Population", setdiff(names(wald_combined), "Population"))]
  }
  
  # 5. Combine overall_summary (AAPC)
  overall_summary_combined <- do.call(rbind, lapply(seq_along(results_list), function(i) {
    df <- results_list[[i]]$overall_summary
    if (!is.null(df) && nrow(df) > 0) {
      df$Population <- population_labels[i]
      df[, c("Population", setdiff(names(df), "Population"))]
    } else {
      NULL
    }
  }))
  
  return(list(
    segment_results = segment_combined,
    bic_candidates = bic_combined,
    change_points = change_points_combined,
    wald_adjacent_slopes = wald_combined,
    overall_summary = overall_summary_combined
  ))
}

# Create list of all results
all_results <- list(
  segmented_results,
  PDO_segmented_results
)

# Create corresponding population labels
population_labels <- c(
  "All (First diagnosed in Australia)",
  "Previously diagnosed overseas"
)

# Combine all results
combined_results <- combine_segmented_results(all_results, population_labels)

# Access the combined datasets
combined_results$segment_results
combined_results$bic_candidates
combined_results$change_points
combined_results$wald_adjacent_slopes
combined_results$overall_summary


# If file not already created make sure it is not overwritten
if (!file.exists(file.path(resultsFolder, "combined_each_slope_results_20260707.csv"))) {
write.csv(combined_results$segment_results, file.path(resultsFolder, "combined_each_slope_results_20260707.csv"), row.names=TRUE)
}
if (!file.exists(file.path(resultsFolder, "combined_aapc_each_results_20260707.csv"))) {
write.csv(combined_results$overall_summary, file.path(resultsFolder, "combined_aapc_each_results_20260707.csv"), row.names=TRUE)
}
if (!file.exists(file.path(resultsFolder, "combined_bic_results_20260707.csv"))) {
write.csv(combined_results$bic_candidates, file.path(resultsFolder, "combined_bic_results_20260707.csv"), row.names=TRUE)
}

if (!file.exists(file.path(resultsFolder, "combined_change_points_results_20260707.csv"))) {
write.csv(combined_results$change_points, file.path(resultsFolder, "combined_change_points_results_20260707.csv"), row.names=TRUE)
}
if (!file.exists(file.path(resultsFolder, "combined_wald_adjacent_slopes_results_20260707.csv"))) {
write.csv(combined_results$wald_adjacent_slopes, file.path(resultsFolder, "combined_wald_adjacent_slopes_results_20260707.csv"), row.names=TRUE)
}


#### annual trend for unknown variable ####
#au_birth unknown
aubirth_unknown <-  hiv_data %>%
  filter(
    year_diag >= 2001,
    year_diag <= 2024,
    previ_diag_overseas == 0,
    is.na(au_birth)
  ) %>%
  group_by(year_diag) %>%
  summarise(
    hiv_cases = n(),
    .groups = "drop"
  )

    
```



```{r missingness calculation}

#### Missingness summary for Limitations / Methods ####
library(dplyr)
library(tidyr)

# Restrict to the analytic sample: first-ever diagnoses, 2001-2024
miss_data <- hiv_data %>%
  filter(year_diag >= 2001,
         year_diag <= 2024,
         previ_diag_overseas == 0)

vars_to_check <- c("au_birth", "age_hiv_notif", "hiv_exposure_risk","sex",
                   "cd4_count_diag", "phn_res")

# Reusable summariser: % missing for each variable in a given subset
missing_summary <- function(df, period_label) {
  df %>%
    summarise(across(all_of(vars_to_check),
                     list(
                       n_missing = ~ sum(is.na(.x)),
                       n_total   = ~ length(.x),
                       pct       = ~ round(100 * mean(is.na(.x)), 2)
                     ),
                     .names = "{.col}__{.fn}")) %>%
    pivot_longer(everything(),
                 names_to = c("variable", "stat"),
                 names_sep = "__",
                 values_to = "value") %>%
    pivot_wider(names_from = stat, values_from = value) %>%
    mutate(period = period_label,
           summary = paste0(n_missing, " / ", n_total, " (", pct, "%)"))
}

# Full period + the two sub-periods RG asked for
missing_by_period <- bind_rows(
  missing_summary(miss_data, "2001-2024"),
  missing_summary(filter(miss_data, year_diag >= 2001, year_diag <= 2010), "2001-2010"),
  missing_summary(filter(miss_data, year_diag >= 2015, year_diag <= 2024), "2015-2024")
)

# Wide view: one row per variable, one column per period (easy to read into the sentence)
missing_wide <- missing_by_period %>%
  dplyr::select(variable, period, summary) %>%
  pivot_wider(names_from = period, values_from = summary)

print(missing_by_period)
print(missing_wide)




```

```{r timeseries plot for phn}

library(ggplot2)
library(dplyr)
library(ggthemes)
library(scales) 
library(extrafont)
library(broom)

#### set time period for rate calculation ####
# Rates are per 100,000 population. time_period = 12 expresses them as a monthly
# average (annual count / 12); set time_period = 1 for annual rates. National
# denominators from ABS; PHN denominators built by allocating SA2 population
# to PHNs via the concordance ratios (RATIO_FROM_TO).
time_period <- 12   # 12 = monthly average rate per 100,000; 1 = annual rate per 100,000
period_label <- ifelse(time_period == 12, "per month", 
                ifelse(time_period == 1, "per year",
                       sprintf("per 1/%d year", time_period)))

#### Rate of HIV diagnoses per 100,000 population by year ####
# Step 1: Read national population data
au_pop_yearly <- read_csv(file.path(dataFolder, "AU_pop_size_2001_2024.csv"),
  col_types = cols(
    Country = col_character(),
    pop_size_2001 = col_double(),
    pop_size_2002 = col_double(),
    pop_size_2003 = col_double(),
    pop_size_2004 = col_double(),
    pop_size_2005 = col_double(),
    pop_size_2006 = col_double(),
    pop_size_2007 = col_double(),
    pop_size_2008 = col_double(),
    pop_size_2009 = col_double(),
    pop_size_2010 = col_double(),
    pop_size_2011 = col_double(),
    pop_size_2012 = col_double(),
    pop_size_2013 = col_double(),
    pop_size_2014 = col_double(),
    pop_size_2015 = col_double(),
    pop_size_2016 = col_double(),
    pop_size_2017 = col_double(),
    pop_size_2018 = col_double(),
    pop_size_2019 = col_double(),
    pop_size_2020 = col_double(),
    pop_size_2021 = col_double(),
    pop_size_2022 = col_double(),
    pop_size_2023 = col_double(),
    pop_size_2024 = col_double()
  )
) %>%
  pivot_longer(
    cols = starts_with("pop_size_"),
    names_to = "year",
    values_to = "total_pop",
    names_prefix = "pop_size_"
  ) %>%
  mutate(year = as.numeric(year)) %>%
  filter(year >= 2001, year <= 2024)

# Step 2: Summarise HIV notifications nationally
hiv_data_pdo <- hiv_data %>%
  filter(
    year_diag >= 2001,
    year_diag <= 2024,
    !is.na(year_diag),
    previ_diag_overseas == 1# advanced_hiv_diag_flag cd4_50_hiv_diag_flag late_hiv_diag_flag previ_diag_overseas
  )

hiv_data_fed <- hiv_data %>%
  filter(
    year_diag >= 2001,
    year_diag <= 2024,
    !is.na(year_diag),
    previ_diag_overseas == 0 # advanced_hiv_diag_flag cd4_50_hiv_diag_flag late_hiv_diag_flag previ_diag_overseas
  )

hiv_cases_au_pdo <- hiv_data_pdo %>%
  group_by(year_diag) %>%
  summarise(
    hiv_cases = n(),
    .groups = "drop"
  ) %>%
  rename(year = year_diag)

hiv_cases_au_fed <- hiv_data_fed %>%
  group_by(year_diag) %>%
  summarise(
    hiv_cases = n(),
    .groups = "drop"
  ) %>%
  rename(year = year_diag)

# Step 3: Join and calculate national notification rate
hiv_au_rate_pdo <- au_pop_yearly %>%
  left_join(hiv_cases_au_pdo, by = "year") %>%
  mutate(
    hiv_cases = ifelse(is.na(hiv_cases), 0, hiv_cases),
    notif_rate_per_100k = (hiv_cases / total_pop) * 100000/time_period
  )

hiv_au_rate_fed <- au_pop_yearly %>%
  left_join(hiv_cases_au_fed, by = "year") %>%
  mutate(
    hiv_cases = ifelse(is.na(hiv_cases), 0, hiv_cases),
    notif_rate_per_100k = (hiv_cases / total_pop) * 100000/time_period
  )

# View the results
print(hiv_au_rate_pdo)

# View the results
print(hiv_au_rate_fed)

## Apply to PDO, FED, and LATE sets
nat_trend_pdo  <- fit_national_rate_trend(hiv_au_rate_pdo)
nat_trend_fed  <- fit_national_rate_trend(hiv_au_rate_fed)

nat_rate_trend_final <- tidy_nat_rate_trends(
  PDO  = nat_trend_pdo,
  FED  = nat_trend_fed
)

nat_rate_trend_final


if (!file.exists(file.path(resultsFolder, "nat_rate_trend_2001_2024_20260707.csv"))) {
write.csv(nat_rate_trend_final, file.path(resultsFolder, "nat_rate_trend_2001_2024_20260707.csv"), row.names=FALSE)
}

if (!file.exists(file.path(resultsFolder, "hiv_au_rate_fed_2001_2024_20260707.csv"))) {
write.csv(hiv_au_rate_fed, file.path(resultsFolder, "hiv_au_rate_fed_2001_2024_20260707.csv"), row.names=FALSE)
}


if (!file.exists(file.path(resultsFolder, "hiv_au_rate_pdo_2001_2024_20260707.csv"))) {
write.csv(hiv_au_rate_pdo, file.path(resultsFolder, "hiv_au_rate_pdo_2001_2024_20260707.csv"), row.names=FALSE)
}

#### Rate of HIV diagnoses per 100,000 population by PHN and year ####
# Read in the lookup tables
pop_sa2_yearly <- read_csv(file.path(dataFolder, "SA2_pop_size_2001_2024.csv"), col_types = cols(
  SA2_CODE21 = col_character(),
  SA2_NAME21 = col_character(),
  pop_size_2001 = col_double(),
  pop_size_2002 = col_double(),
  pop_size_2003 = col_double(),
  pop_size_2004 = col_double(),
  pop_size_2005 = col_double(),
  pop_size_2006 = col_double(),
  pop_size_2007 = col_double(),
  pop_size_2008 = col_double(),
  pop_size_2009 = col_double(),
  pop_size_2010 = col_double(),
  pop_size_2011 = col_double(),
  pop_size_2012 = col_double(),
  pop_size_2013 = col_double(),
  pop_size_2014 = col_double(),
  pop_size_2015 = col_double(),
  pop_size_2016 = col_double(),
  pop_size_2017 = col_double(),
  pop_size_2018 = col_double(),
  pop_size_2019 = col_double(),
  pop_size_2020 = col_double(),
  pop_size_2021 = col_double(),
  pop_size_2022 = col_double(),
  pop_size_2023 = col_double(),
  pop_size_2024 = col_double()
))

phn_lookup <- read_csv(file.path(dataFolder, "phn_sa2_20250901.csv"), col_types = cols_only(
  SA2_NAME21 = col_character(),
  SA2_CODE21 = col_character(),
  PHN_CODE = col_character(),
  PHN_NAME = col_character(),
  RATIO_FROM_TO = col_double()
))

library(ggtext)  # for colored axis labels

#### population data by PHN and year ####
# Transform population data from wide to long format
pop_sa2_long <- pop_sa2_yearly %>%
  pivot_longer(
    cols = starts_with("pop_size_"),
    names_to = "year",
    values_to = "pop_size",
    names_prefix = "pop_size_"
  ) %>%
  mutate(year = as.numeric(year)) %>%
  filter(year >= 2001, year <= 2024)

# Calculate PHN population for each year
phn_pop_yearly <- pop_sa2_long %>%
  inner_join(phn_lookup, by = "SA2_CODE21",relationship = "many-to-many") %>%
  mutate(
    pop_alloc = pop_size * RATIO_FROM_TO
  ) %>%
  filter(!is.na(PHN_CODE)) %>%
  group_by(PHN_CODE, PHN_NAME, year) %>%
  summarise(
    total_pop = sum(pop_alloc, na.rm = TRUE),
    .groups = "drop"
  )

yearly_totals <- phn_pop_yearly %>%
  group_by(year) %>%
  summarise(
    total_australia_pop = sum(total_pop, na.rm = TRUE),
    num_phns = n_distinct(PHN_CODE),
    .groups = "drop"
  ) %>%
  arrange(year)

#### PHN case rates + trend (FED and PDO) ####
hiv_case_rates_fed <- compute_phn_case_rates(hiv_data, 0, phn_pop_yearly, time_period)
hiv_case_rates_pdo <- compute_phn_case_rates(hiv_data, 1, phn_pop_yearly, time_period)

phn_case_trend_final_fed <- compute_phn_trend(hiv_case_rates_fed)
phn_case_trend_final_pdo <- compute_phn_trend(hiv_case_rates_pdo)

if (!file.exists(file.path(resultsFolder, "phn_case_trend_FED_2001_2024_20260707.csv"))) {
write.csv(phn_case_trend_final_fed, file.path(resultsFolder, "phn_case_trend_FED_2001_2024_20260707.csv"), row.names=FALSE)
}
if (!file.exists(file.path(resultsFolder, "phn_case_trend_PDO_2001_2024_20260707.csv"))) {
write.csv(phn_case_trend_final_pdo, file.path(resultsFolder, "phn_case_trend_PDO_2001_2024_20260707.csv"), row.names=FALSE)
}

#### Heatmaps ####
# Define breaks and colors for HIV case rates per 100k
# Adjust these based on actual data range

common_breaks <- c(0, 0.25,0.5,0.75, 1,1.25, 1.5, 1.75)
common_colors <- c("#FFFFCC", "#C7E9B4", "#7FCDBB", "#41B6C4",
                   "#1D91C0", "#225EA8", "#0C2C84", "#081D58")
common_labels <- c("0·0", "","0·5","", "1·0","", "1·5", "")
fed <- make_phn_heatmap(hiv_case_rates_fed,
        "A. Monthly average notification rates, First ever diagnosed in Australia")
pdo <- make_phn_heatmap(hiv_case_rates_pdo,
        "B. Monthly average notification rates, Previously diagnosed overseas")
print(fed)
print(pdo)

library(patchwork)
combined_plot_phn_hiv_rates_2001_24 <-
  fed / pdo +
  plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")


####save figure for combined plot####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 40, 
    height = 32, units = "cm", dpi = 600,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}

if (!file.exists(file.path(figFolder, "combined_plot_phn_hiv_rates_2001_24_20260707.png"))) {
    SaveFigure(figFolder,"combined_plot_phn_hiv_rates_2001_24_20260707", combined_plot_phn_hiv_rates_2001_24)
}


#### small summaries + CSV (FED & PDO) ####
phn_rate_diff_fed <- hiv_case_rates_fed %>%
  filter(year %in% c(2001, 2024)) %>%
  dplyr::select(PHN_CODE, PHN_NAME, year, case_rate_per_100k) %>%
  pivot_wider(names_from = year, values_from = case_rate_per_100k, names_prefix = "rate_") %>%
  mutate(rate_diff_2024_minus_2001 = rate_2024 - rate_2001)

phn_rate_diff_pdo <- hiv_case_rates_pdo %>%
  filter(year %in% c(2001, 2024)) %>%
  dplyr::select(PHN_CODE, PHN_NAME, year, case_rate_per_100k) %>%
  pivot_wider(names_from = year, values_from = case_rate_per_100k, names_prefix = "rate_") %>%
  mutate(rate_diff_2024_minus_2001 = rate_2024 - rate_2001)


if (!file.exists(file.path(resultsFolder, "phn_rate_diff_fed_2001_2024_20260707.csv"))) {
write.csv(phn_rate_diff_fed, file.path(resultsFolder, "phn_rate_diff_fed_2001_2024_20260707.csv"), row.names=FALSE)
}

if (!file.exists(file.path(resultsFolder, "phn_hiv_case_rates_diff_pdo_2001_2024_20260707.csv"))) {
write.csv(phn_rate_diff_pdo, file.path(resultsFolder, "phn_hiv_case_rates_diff_pdo_2001_2024_20260707.csv"), row.names=FALSE)
}

#### Gini coefficient plot ####
# Geographic inequality in PHN notification rates, summarised by the Gini
# coefficient (0 = evenly distributed, 1 = maximally concentrated) and the
# Lorenz curve. Gini is population-weighted (weighted_gini); PHNs ordered by
# rate.
library(dplyr)
library(ggplot2)
library(scales)


phn_gini_trend <- hiv_case_rates_fed %>%
  group_by(year) %>%
  summarise(
    gini_weighted = weighted_gini(case_rate_per_100k, total_pop),
    mean_rate     = mean(case_rate_per_100k, na.rm = TRUE),
    sd_rate       = sd(case_rate_per_100k, na.rm = TRUE),
    .groups = "drop"
  )

phn_gini_overall_trend <- lm(gini_weighted ~ year, data = phn_gini_trend)
summary(phn_gini_overall_trend)

gini_time_plot <- ggplot(phn_gini_trend, aes(x = year, y = gini_weighted)) +
  geom_line(color = "#1f77b4", linewidth = 1) +
  geom_point(color = "#1f77b4", size = 2) +
  scale_x_continuous(breaks = seq(2001, 2024, 3)) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, 0.25),
                     labels = c("0·00", "0·25", "0·50",  "0·75", "1·00"),
                     expand = c(0, 0)) +   # y-axis starts at 0
  labs(
    title = "",
    y = "Gini coefficient",
    x = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),     # remove major gridlines
    panel.grid.minor = element_blank(),     # remove minor gridlines
    axis.line = element_line(color = "black"),  # add black x & y axes
    axis.ticks = element_line(color = "black"), # add tick marks
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", size = 12)
  )


#Run the lorenz plot 
lorenz_2001 <- plot_phn_lorenz(hiv_case_rates_fed, phn_map, year_selected = 2001,
                               label_left = c("Central and Eastern Sydney"),
                               label_lefttop = c("North Western Melbourne", "South Eastern Melbourne"),
                               label_right = c(),
                               sub_title = "A. Notifications (2001); Gini: 0·45")
lorenz_2001$plot
lorenz_2024 <- plot_phn_lorenz(hiv_case_rates_fed, phn_map, year_selected = 2024,
                               label_left = c("Central and Eastern Sydney"),
                               label_lefttop = c("North Western Melbourne", "Brisbane North"),
                               label_right = c(),
                               sub_title = "B. Notifications (2024); Gini: 0·26")
lorenz_2024$plot
  
library(patchwork)

# Combine plots side-by-side with one legend
combined_lorenz <- lorenz_2001$plot +
                   lorenz_2024$plot +
                   plot_layout(guides = "collect") & theme(legend.position = "bottom")

# Show combined plot
combined_lorenz

#### Save figures ####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 20, 
    height = 12.5, units = "cm", dpi = 600,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}

# If figure not already created make sure it is not overwritten
if (!file.exists(file.path(figFolder, "combined_lorenz_plot_2001_2024_20260707.png"))) {
  SaveFigure(figFolder,"combined_lorenz_plot_2001_2024_20260707", combined_lorenz)
}

if (!file.exists(file.path(figFolder, "gini_time_plot_2001_2024_20260707.png"))) {
  SaveFigure(figFolder,"gini_time_plot_2001_2024_20260707", gini_time_plot)
}


```